---
title: "Analysis of movies"
output: html_document
---
<!-- --- -->
<!-- title: "Session 2: Homework 1" -->
<!-- author: "Study Group 12" -->
<!-- date: "`r Sys.Date()`" -->
<!-- output: -->
<!--   html_document: -->
<!--     theme: flatly -->
<!--     highlight: zenburn -->
<!--     number_sections: yes -->
<!--     toc: yes -->
<!--     toc_float: yes -->
<!--     code_folding: show -->
<!-- --- -->


```{r, setup}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(lubridate)
library(fivethirtyeight)
library(here)
library(ggthemes) 
library(ggrepel) # to avoid overlapping legends
library(skimr)
library(janitor)
library(vroom) 
library(tidyquant)
```

# Analysis of movies- IMDB dataset

We will look at a subset sample of movies, taken from the [Kaggle IMDB 5000 movie dataset](https://www.kaggle.com/carolzhangdc/imdb-5000-movie-dataset)

  
```{r,load_movies, warning=FALSE, message=FALSE}

#loading dataset movies using read_csv
movies <- read_csv(here::here("data", "movies.csv"))
glimpse(movies)

```

We used the skim command to check whether the dataset has missing values and duplicates. The output of the skim function revealed that our dataset does not contain missing values but that certain variables (title, genre and director) do have duplicates. 

```{r, missing_and_duplicates}
skim(movies)
```

Table with the count of movies by genre, ranked in descending order:

```{r count_movies_by_genre}

#Count of movies grouped by genre
movies_by_genre <- movies %>%
  group_by(genre) %>%
  summarise(count_of_movies = n()) %>%
  arrange(desc(count_of_movies))

movies_by_genre

```

Table showing the return on budget grouped by genre:

```{r return_on_budget}

#Calculating the return on budget by genre which is the average gross earning/average budget 
return_on_movie_budgets <- movies %>%
  group_by(genre) %>%
  summarise(average_gross_earning = mean(gross), average_budget = mean(budget)) %>%
  mutate(return_on_budget = average_gross_earning/average_budget) %>%
  arrange(desc(return_on_budget))

return_on_movie_budgets

```

Table showing the summary statitics for revenue for 15 movie directors:

```{r, top_15_directors}

#Revenue statitics for top 15 movie directors
top_15_directors <- movies %>%
  group_by(director) %>%
  summarise(sum_revenue = sum(gross), mean_revenue = mean(gross), median_revenue = median(gross), std_revenue = sd(gross)) %>%
  arrange(desc(sum_revenue)) %>%
  head(15)

top_15_directors

```

Distribution of ratings by genre:

```{r, ratings_by_genre}

# Summarising data grouped by genre 
ratings_by_genre <- movies %>%
  group_by(genre) %>%
  summarise(mean_rating = mean(rating), min_rating = min(rating), max_rating = max(rating), median_rating = median(rating), std_rating = sd(rating))

ratings_by_genre

# Creating a density plot of the spread of ratings for various movie genres
ggplot(movies, aes(x=rating)) +
  geom_density() +
  facet_wrap(~genre)+
  labs(title = "Density plots of the distribution of ratings for different genres",
    y = "Density",
    x = "Rating")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

# Creating a box plot showing the spread of ratings for various movie genres
ggplot(movies, aes(x=genre, y=rating)) +
  geom_boxplot() +
   labs(title = "Box plot of the distribution of ratings for different genres",
    y = "Rating ",
    x = "Genres")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), axis.text.x = element_text(angle = 90)
  )

```

## Use `ggplot` to answer the following
  
We created a scatterplot between gross revenue and cast Facebook likes and calculated the correlation between the two variables. A weak positive correlation (0.213) between `gross` and `cast_facebook_likes` suggests suggests that the two have no multicollinearity. On the X axis we have mapped `cast_facebook_likes` and on the Y axis we have mapped the variable `gross`. 
  
```{r, gross_on_fblikes}

#Scatterplot between Gross Revenue and Cast Facebook Likes
ggplot(movies, aes(x=cast_facebook_likes, y=gross)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth()+
   labs(title = "Scatter plot between Gross Revenue and Cast Facebook Likes",
    y = "Gross Revenue",
    x = "Cast Facebook Likes")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

#calculating the correlation between the two variables
cor(movies$gross, movies$cast_facebook_likes)

```

Next, we created a scatterplot between `gross` revenue and `budget` and calculated the correlation between the two variables.Correlation of 0.641 indicates that budget might be a better predictor of gross revenue when compared with cast Facebook likes but further analysis will be needed. 

```{r, gross_on_budget}

#Scatterplot of gross revenue vs budget
ggplot(movies, aes(x=budget, y=gross)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth()+
   labs(title = "Scatter plot between Gross Revenue and Budget",
    y = "Gross Revenue",
    x = "Budget")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

#calculating the correlation between the two variables
cor(movies$gross, movies$budget)


```
  
Finally, we created a scatterplot between gross revenue and ratings, faceted by genre and calculated the correlation between the two variables. A weak positive correlation (0.269) between `gross` and `rating` suggests that the two have no multicollinearity. After perfroming a linear regression (significance level <0.001) we concluded that the rating is a good predictor of movie revenue at the box office. 
  
```{r, gross_on_rating}

#Scatterplots of gross revenue vs rating for various genres
ggplot(movies, aes(x=rating, y=gross)) +
  geom_point(alpha = 0.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~genre) +
  geom_smooth()+
  labs(title = "Scatter plot between Gross Revenue and Rating",
    y = "Gross Revenue",
    x = "Rating")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

#calculating the correlation between the two variables
cor(movies$gross, movies$rating)
model<-lm(gross~rating,data=movies)
summary(model)

```

