---
title: "Gapminder Revisited"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
```

# Gapminder revisited

Importing and cleaning data:

```{r, get_data, cache=TRUE}

# load gapminder HIV data
hiv <- read_csv(here::here("data","adults_with_hiv_percent_age_15_49.csv"))
life_expectancy <- read_csv(here::here("data","life_expectancy_years.csv"))

# get World bank data using wbstats
indicators <- c("SP.DYN.TFRT.IN","SE.PRM.NENR", "SH.DYN.MORT", "NY.GDP.PCAP.KD")


library(wbstats)

worldbank_data <- wb_data(country="countries_only", #countries only- no aggregates like Latin America, Europe, etc.
                          indicator = indicators, 
                          start_date = 1960, 
                          end_date = 2016)

# get a dataframe of information regarding countries, indicators, sources, regions, indicator topics, lending types, income levels,  from the World Bank API 
countries <-  wbstats::wb_cachelist$countries

```

We first transformed the life expectancy and HIV data into long format and performed an inner join on the year and country column to remove observations that were not common between the two datasets. We then did an inner join of the combined data (HIV and life expectancy) with the world bank data and added the regions column from the countries dataset. We named the combined dataset `wb_lifeExp_hiv`. Finally, we created an extended version of the world bank dataset by adding the regions column from `countries` onto it. 

```{r, tidy_economic_data}

# putting the life expectancy data into long format
life_expectancy <- life_expectancy %>% 
  pivot_longer(cols="1800":"2100",  values_to= "Life_Exp", names_to = "Year")

# putting the hiv data into long format
hiv <- hiv %>% 
  pivot_longer(cols="1979":"2011",  values_to= "HIV", names_to = "Year")


#extracting the observations (based on year of observation) from the life expectancy data for which there are observations in the HIV dataset
life_expectancy_cleaned <- life_expectancy %>% 
  mutate(Year= as.character(Year))


#joining the life exp and HIV datasets
life_exp_hiv <-  
  inner_join(life_expectancy_cleaned, hiv, by= c("Year", "country" ))


worldbank_data <- 
  worldbank_data %>% 
  rename(Year="date") %>% 
  mutate(Year= as.character(Year))

#joining all three datasets and adding in the region column from the countries dataset
wb_lifeExp_hiv <- inner_join(worldbank_data, life_exp_hiv, by=c("country", "Year"))

wb_lifeExp_hiv <- inner_join(wb_lifeExp_hiv, countries %>% select(region, country), by=c("country"))

world_bank_extended <- inner_join(worldbank_data, countries %>% select(region, country), by=c("country"))
               

```

Relationship between life expectancy and HIV prevalence:

```{r, life expectancy and HIV prevalence, fig2, fig.width = 12, fig.height=8}


#plotting the relationship between life expectancy and HIV prevalence 
ggplot(wb_lifeExp_hiv, aes(x= HIV, y=Life_Exp)) +
  geom_point(size=0.08, alpha=0.3)+
  geom_smooth(method = "lm")+
  scale_x_log10()+
  labs (
    title = "Life Expectancy  vs HIV Prevalence by Region",
    x= "HIV Prevalence per 100 population",
    y="Life Expectancy"
    
  )+
  theme_clean()+
  facet_wrap(~region)


```


Relationship between fertility rate and GDP per capita:

From the graph below we can clearly see that as GDP per capita increases, the fertility rate decreases as women who are employed tend to have fewer children, slightly later in their lives. Moreover, people in rich countries tend to have lesser children because children need a higher level of education to suceed in richer countries and therefore prove to be more expensive.

```{r, fertility_gdp_per_cap, fig3, fig.width = 12, fig.height=8}

#plotting the relationship between life expectancy and HIV prevalence 
ggplot(world_bank_extended, aes(x= NY.GDP.PCAP.KD/1000, y=SP.DYN.TFRT.IN)) +
  geom_point(size=0.08, alpha=0.3)+
  geom_smooth(method='lm')+
  scale_x_log10()+
  labs (
    title = "Relationship between fertility rate and GDP per capita",
    x="GDP per capita (in ,000 constant 2010 USD)",
    y="Fertility Rate",
    caption=" Source: World Bank"
  )+
  theme_clean()+
  facet_wrap(~region)+
  scale_x_continuous(labels=scales::dollar_format())

```

Regions that have observations with missing HIV data:

From the graph we can see that in this dataset, Europe and Central Asia have the most observations with missing data for HIV. 

```{r, observations with missing HIV data, fig3, fig.width = 12, fig.height=8}


wb_lifeExp_hiv %>% 
  #filtering for observations with missing values
  group_by(region) %>% 
  filter(is.na(HIV)) %>% 
  summarise(count=n()) %>% 
  ggplot(aes(x=fct_reorder(region, count), y= count))+
  geom_col()+
  coord_flip()+
  labs (
    title = "Europe and central asia have the most observations with missing HIV data!",
    y= "Number of observations with missing HIV data"
  )+
  theme_clean()+
  theme(
    axis.title.y=element_blank()
  )


```


```{r, mortality rate}

worldbank_data <- worldbank_data%>%
  rename(date="Year")
  
#Add region information using the countries dataframe, joined on country column
mortality <- left_join(worldbank_data, countries[c("country", "region")], by = "country")

#By region
#Create mortality rate table by year and region by averaging
mortality_regions <- na.exclude(mortality[c("region", "country", "date", "SH.DYN.MORT")])
mortality_regions <- mortality_regions %>%
  group_by(date, region) %>%
  summarise(mortality = mean(SH.DYN.MORT)) %>%
  arrange(region, date)

#Create table with start mortality rate by year and region
mortality_regions_start <- mortality_regions[!duplicated(mortality_regions$region, mortality_regions$date),]
mortality_regions_start <- mortality_regions_start[c("region", "mortality")] %>%
  arrange(region)

#Create table with end mortality rate by year and region
mortality_regions <- mortality_regions %>%
  arrange(desc(date), desc(region))
mortality_regions_stop <- mortality_regions[!duplicated(mortality_regions$region, mortality_regions$date),]
mortality_regions_stop <- mortality_regions_stop[c("region", "mortality")] %>%
  arrange(region)

#Create table with change in mortality rate by region
mortality_change_regions <- data.frame(region = mortality_regions_start$region, mortality_change = (mortality_regions_stop[c("mortality")] - mortality_regions_start[c("mortality")]) / mortality_regions_start[c("mortality")])

#By country
#Create mortality rate table by year and country by averaging
mortality_countries <- na.exclude(mortality[c("region", "country", "date", "SH.DYN.MORT")])
mortality_countries <- mortality_countries %>%
  group_by(date, country) %>%
  summarise(mortality = mean(SH.DYN.MORT)) %>%
  arrange(country, date)

#Create table with start mortality rate by year and country
mortality_countries_start <- mortality_countries[!duplicated(mortality_countries$country, mortality_countries$date),]
mortality_countries_start <- mortality_countries_start[c("country", "mortality")] %>%
  arrange(country)

#Create table with end mortality rate by year and country
mortality_countries <- mortality_countries %>%
  arrange(desc(date), desc(country))
mortality_countries_stop <- mortality_countries[!duplicated(mortality_countries$country, mortality_countries$date),]
mortality_countries_stop <- mortality_countries_stop[c("country", "mortality")] %>%
  arrange(country)

#Create table with change in mortality rate by country
mortality_change_countries <- data.frame(country = mortality_countries_start$country, mortality_change = (-mortality_countries_stop[c("mortality")] + mortality_countries_start[c("mortality")]) / mortality_countries_start[c("mortality")])

mortality_change_countries <- mortality_change_countries %>% 
  mutate(mortality= mortality*100) %>% 
  rename(Percentage_Change_in_mortality_rate= "mortality") %>% 
  inner_join(countries %>% select(region, country), by=c("country"))


#display top 5 and bottom 5 countries for various regions
paste("top 5 countries that have seen greatest improvement in Sub-Saharan Africa")

mortality_change_countries %>% 
  filter(region== "Sub-Saharan Africa") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in Sub-Saharan Africa")

mortality_change_countries %>% 
  filter(region== "Sub-Saharan Africa") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)

paste("top 5 countries that have seen greatest improvement in Middle East & North Africa")

mortality_change_countries %>% 
  filter(region== "Middle East & North Africa") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in Middle East & North Africa")

mortality_change_countries %>% 
  filter(region== "Middle East & North Africa") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)

paste("top 5 countries that have seen greatest improvement in Europe & Central Asia")

mortality_change_countries %>% 
  filter(region== "Europe & Central Asia") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in Europe & Central Asia")

mortality_change_countries %>% 
  filter(region== "Europe & Central Asia") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)

paste("top 5 countries that have seen greatest improvement in Latin America & Caribbean")

mortality_change_countries %>% 
  filter(region== "Latin America & Caribbean") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in Latin America & Caribbean")

mortality_change_countries %>% 
  filter(region== "Latin America & Caribbean") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)

paste("top 5 countries that have seen greatest improvement in South Asia")

mortality_change_countries %>% 
  filter(region== "South Asia") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in South Asia")

mortality_change_countries %>% 
  filter(region== "South Asia") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)

paste("top 5 countries that have seen greatest improvement in East Asia & Pacific")

mortality_change_countries %>% 
  filter(region== "East Asia & Pacific") %>% 
  arrange(desc(Percentage_Change_in_mortality_rate)) %>% 
  head(5)

paste("5 countries that have seen least improvement in East Asia & Pacific")

mortality_change_countries %>% 
  filter(region== "East Asia & Pacific") %>% 
  arrange(Percentage_Change_in_mortality_rate) %>% 
  head(5)



  

```
