---
title: "Beer, wine, and Spirits"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(lubridate)
library(fivethirtyeight)
library(here)
library(ggthemes) 
library(ggrepel) # to avoid overlapping legends
library(skimr)
library(janitor)
library(vroom) 
library(tidyquant)
```


# Where Do People Drink The Most Beer, Wine And Spirits?

Back in 2014, [fivethiryeight.com](https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/) published an article on alchohol consumption in different countries. The data `drinks` is available as part of the `fivethirtyeight` package. Make sure you have installed the `fivethirtyeight` package before proceeding.


```{r, load_alcohol_data}
library(fivethirtyeight)
data(drinks)

alcohol_direct <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/alcohol-consumption/drinks.csv")

```


We checked column types and missing values using the *skim* function and observed that all the columns are of type "numeric", except the "country" column, which is of type "character". We also observed that there are no missing values.

```{r glimpse_skim_data}
skim(alcohol_direct)
```


Plot that shows the top 25 beer consuming countries:

```{r beer_plot}
#selecting the top 25 beer drinking countries from the dataset
top_25_b <- drinks %>%
  arrange(desc(beer_servings)) %>%
  select(country, beer_servings) %>%
  head(25)
  
  
#plotting the top 25 beer drinking countries using a bar chart
ggplot(top_25_b, aes(x=fct_reorder(country, beer_servings), y=beer_servings))+
  geom_col()+
  coord_flip()+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  ) +
  labs(title = "Top 25 beer consuming countries",
    y = "Beer Servings consumed per person",
    x = "Country")


```

Plot that shows the top 25 wine consuming countries:

```{r wine_plot}

#selecting the top 25 wine drinking countries from the dataset
top_25_w <- drinks %>%
  arrange(desc(wine_servings)) %>%
  select(country, wine_servings) %>%
  top_n(25)
  

#plotting the top 25 wine drinking countries using a bar chart
ggplot(top_25_w, aes(x=fct_reorder(country, wine_servings),  y=wine_servings))+
  geom_col()+
  coord_flip()+
    labs(title = "Top 25 wine consuming countries",
    y = "Wine Servings consumed per person",
    x = "Country")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )


```

Plot that shows the top 25 spirit consuming countries:

```{r spirit_plot}

#selecting the top 25 spirit drinking countries from the dataset
top_25_s <- drinks %>%
  arrange(desc(spirit_servings)) %>%
  select(country, spirit_servings) %>%
  top_n(25)
  
#plotting the top 25 spirit drinking countries using a bar chart
ggplot(top_25_s, aes(x=fct_reorder(country, spirit_servings), y=spirit_servings))+
  geom_col()+
  coord_flip()+
    labs(title = "Top 25 spirit consuming countries",
    y = "Spirit Servings consumed per person",
    x = "Country")+
    theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

```

From the graphs above we can see that on average Namibians consume the most beer, the French consume the most wine and Grenanda consumes the most spirits per person per year. Differences in culture, availability of ingredients (barley,grapes, corn etc) in the country and regulations on liquor sale might be possible reasons for the differences in consumption of wine, beer and spirits across countries. 